SignalHsmm - a novel semi-Markov model of eukaryotic signal peptides

### 2015-06-20 paper.tex
improvements from PM + resized tables and figures to fit on 10 pages

\title{signalHsmm - a novel semi-Markov model of eukaryotic signal peptides}

@@

\keywords{signal petide prediction; n-gram; hidden semi-Markov models}

\begin{abstract}
The proper localization of proteins in a cell is essential to maintain their desired function. Information about the protein destination is included within the very protein in the form of short peptides called targeting signals. Ones of them are signal peptides, diverse N-terminal sequences, which are responsible for targeting of proteins to endomembrane system and their export outside the cell. Proteins equipped with signal peptides play crucial roles in metabolism, maintenance of tissue structure, immune response and regulation of other organismal functions.

A common model of classical signal peptides assumes that they start with a positively charged n-region, followed by a hydrophobic h-region and a c-region ended with a cleavage site recognized by a signal peptidase. However, studies of many protein sequences representing the wide range of diversified taxonomic organisms indicate a variability of signal peptides. Therefore, we designed a new, more universal probabilistic model for eukaryotic signal peptides, which includes knowledge about their organization, amino acid composition and variation.

The proposed model is based on hidden semi-Markov models (HSMMs) and use intrinsic knowledge about signal peptides. The big advantage of the algorithm is its extensibility. Using the n-grams (k-mers) we point how the general model can be attuned to yield not only better results, but also more information about signal peptides.

Our model was validated as a signal peptide predictor. It has showed the largest AUC=0.97 in comparison to other software and appeared very stable in the recovery of signal peptides after training even on very small data sets. Thanks to that, our model does not need to be permanently retrained with the continuous expansion of sequence databases. It should be emphasized that our model describes signal peptides from medically significant malaria parasites Plasmodium and their relatives (AUC = 0.92) more accurately than popular programs (0.84).

The web-server of signalHsmm is avaible under address \url{http://smorfland.uni.wroc.pl/signalhsmm}.

@@

\subsection*{Signal peptides}

Proteins of eukaryotes are encoded in nuclear genomes and are synthesized in ribosomes located in the cytosol or bounded by the endoplasmic reticulum. After the translation process, proteins have to be targeted to specific subcellular compartments or exported outside the cell to the extracellular environment. The proper localization of proteins is essential to perform their desired function. Information about protein destination is included within the very protein in the form of short peptides or stretches of amino acid residues called targeting or sorting signals. Ones of them are signal peptides, which are located at the N-terminus of proteins.

Signal peptides are responsible for targeting of proteins via the Sec61 translocation channel~\citep{2007rapoportprotein} to endomembrane system, which includes endoplasmic reticulum, Golgi apparatus and endosomes. Such proteins can stay inside of these compartments, or can be inserted into cellular membranes or exported outside the cell. Proteins equipped with signal peptides constitute a substantial fraction of the whole proteome. They play crucial roles in metabolism ($\beta$ galactosidase, pepsins)~\citep{1991hofmannmutations}, maintenance of tissue structure (collagen)~\citep{2001chanaberrant}, immune response (interferons, interleukins)~\citep{2005zhangalteration} and regulation of other organismal functions (prolactin, glucagon)~\citep{2010huangrole}. Moreover, passing proteins through the endomembrane system is important for their correct folding and posttranslational modification such as glycosylation and phosphorylation.

Despite the low sequence homology between signal peptides~\citep{1999ladungaphysean}, some general architecture were proposed~\citep{1994izardsignal, 2013vossmechanism} - Fig. \ref{fig:sparch}. It is assumed that signal peptides start with a positively charged sequence of amino acid residues, called the n-region with the length of about 5-8 residues. They probably enforce a proper topology on the polypeptide during translocation through membrane based on the positive-inside rule~\citep{1988vonheijnetopogenic}. The first region is followed by a stretch of hydrophobic amino acids (h-region) with the length of about 8-12 residues. It constitutes a core region of signal peptide and has a tendency to form $\alpha$-helix. The third part of a signal peptide, usually 6 residues long, is a polar, but uncharged c-region ended with a cleavage site recognized by the signal peptidase. The amino acid composition and the length of these regions vary between signal peptides, which influences the efficiency of protein secretion~\citep{2006hegdethe}.

\begin{figure}[ht]\centering
\includegraphics[width=0.55\textwidth]{figures/SP.png}
\caption{The organization of signal peptide. The figure is not drawn to scale.}
\label{fig:sparch}
\end{figure}

During or after translocation of the protein into the lumen of endoplasmic reticulum, the typical signal peptide is cleaved by a signal peptidase~\citep{2002paetzelsignal} and next degraded by specific proteases, whereas the rest (mature) part of protein stays in the lumen or is passed to other compartments. The cleavage site is characterized by a very variable amino acid composition. It typically contains small and neutral residues at -3 and -1 positions~\citep{1994palzkillselection}. The site is, however, absent from some membrane proteins in which the first transmembrane-domain acts both as signal peptides and signal anchor~\citep{1988szczesnaskorupapositive}.

Some data indicate that signal peptides may be universal. It was found, for example, that even bacterial signal peptides targeted correctly transgenic proteins to the plant~\citep{2009moellera} or mammalian secretory system~\citep{2014naganoestablishment}. On the other hand, signal peptides show great variation and the description presented above (Fig. \ref{fig:sparch}) refers to the most ‘typical’ signal peptides. There are also exceptionally long signal peptides, which fulfill more sophisticated roles~\citep{2009hissarchitecture}. For example, the fragment of signal peptide from preprolactin probably takes part in the regulation of prolactin secretion. Other examples are signal peptides of MHC class I, which inhibit activity of NK cells. Interesting functions have signal peptides of viral origin, which are involved in the immune evasion or viral life cycle~\citep{2000kappposttargeting}. Such diverse functions are restricted not only to the long signal peptides. The peptides from midkine, a protein contributing to the tumor progression, contains epitopes recognized by CD4+ T cells~\citep{2013kerzerhothe}. It indicates that the signal peptide may participate in a tumor immunity. The functional significance of these targeting signals makes that the prediction of signal peptide-containing proteins is also an important step in the drug development~\citep{2005zhangalteration, 2012netoadeimproving, 2010moellerwetmilling}.

\subsection*{Signal peptide predicting software}

Although many experimental methods determining the subcellular localization of proteins were devised, they are time consuming and laborious. Therefore, the development of new approaches in the field of computational biology and bioinformatics is desirable. They are not only a good support, complement or alternative for the experimental methods but also enable to understand rules encoding information about protein targeting, which is contained in the predicted targeting signals. However, many of the present predictors disregard full biological information carrying by signal peptides.

Having functional importance and specific features, signal peptides became the subject of many programs to their prediction based on different methods. The state of the art software for predicting the presence of signal peptides often incorporates ‘black-box’ models, as neural networks~\citep{2011petersensignalp}, support vector machines~\citep{2014zhangprediction}, Bayesian networks~\citep{2012zhengsignalbnf} or k-nearest neighbours~\citep{2007shensignall}. Such models neither utilize biological information about signal peptides nor work properly on atypical signal peptides.

The other type of software incorporates into the prediction process theoretical knowledge about the structure of signal peptides. Although these programs do not share innate flaws of ‘black-box’ models, they also demand an improvement. Some of them are based on position matrices or their more abstract variants~\citep{2014zhangprediction, 2004hillerpredisi}. Others (Phobius, Philius and not supported SignalP 3.0) use hidden Markov models (HMMs)~\citep{2004klla, 2008reynoldstransmembrane, 2004bendtsenimproved}, which reflect regions of signal peptides without realizing limitations of this probabilistic framework. However, the HMMs imply geometric distribution for duration of region length. We replicated the rules for extracting regions' boundaries from the first work utilizing HMMs in prediction of signal peptides~\citep{1998nielsenprediction} and found that the mentioned assumption is not in concordance with the reality because every region revealed the length distribution other than geometric (Fig. \ref{fig:reglen}). Moreover, the commonly used rigid region scheme (Fig. \ref{fig:sparch}) does not describe extremely long or short signal peptides. Theoretically, HMMs that describe the atypical signal peptides could be developed to consider also unusual structures, but such probabilistic frameworks have not still been implemented.

\begin{figure}[ht]\centering
\includegraphics[width=0.95\textwidth]{figures/reglen.png}
\caption{A) Distribution of lengths of signal peptides. B) Distribution of the lengths of signal peptide regions. The measure of the length is a number of amino acids in the subsequence. The data was extracted from 2~589 signal peptide sequences derived from UniProt database (see \textbf{Data selection} in \textbf{Methods}).}
\label{fig:reglen}
\end{figure}

Another feature, shared by majority of the signal peptide predicting software, is the orthogonal encoding of amino acids, in which a vector of 20 digits represents every amino acid. This method of encoding, however, does not preserve relationships between amino acids. The distance between residues is the same regardless of the similarity of their physicochemical properties. Such behaviour is especially undesired in signal peptides models because their regions are defined by specific features of amino acid residues and not by the simple occurrence of specific amino acids. In addition to this, such sparse encoding enforces larger data sets, which hinders their management and analysis~\citep{2002linamino}.

All programs used in signal peptide recognition are trained on real protein sequences. Therefore, they succeed in the recognition of peptides similar to those in the learning set, but fail in the case of artificial signal peptides. Such peptides are designed to increase effectiveness of protein secretion~\citep{2010futatsumorisugaisignal}. They are especially important in industrial applications to increase yield of proteins. Therefore, only explicit knowledge about the organization of signal peptides allows creating sequences that will be the most efficient in the export of proteins~\citep{2013ngengineering}. Signal peptides have also an important application in gene therapy. Mimicking the natural mechanism of protein export, artificial signal peptides with tumor epitopes increase the antitumor immune response~\citep{2003heenhanced}. Such epitopes must be properly inserted into a signal peptide without decreasing its secretion properties through disruption of the regional structure. Instead of time-consuming and expensive laboratory experiments, it would be very useful to survey in silico many artificial peptides to select the ones that would fulfil the designed role.

@@

The functionality of a signal peptide depends not on exact sequence of specific amino acids, but on the physicochemical properties of residues in a given region. Henceforth, the usage of raw amino acid sequences is superfluous and introduces unnecessary information. To utilize this property of signal peptide recognition, we clustered amino acids into several groups based on the physicochemical features of residues.

The pre-processed sequences were further analyzed by the heuristic algorithm, which determines borders between three characteristic signal peptide regions, the enhanced version of algorithm presented in~\cite{1998nielsenprediction}. We refined some region recognition criteria to attune the algorithm to less typical signal peptides.

Next, two models are trained to recognize proteins with and without a signal peptide. The first one is a hidden semi-Markov model, in which each of three signal peptide regions is represented by a different hidden state. The additional fourth hidden state represents mature protein. Each state is described by the frequencies of amino acid groups  within that state. The distribution of hidden states durations, the number of amino acids related to each hidden state in signal peptide, is based on the empirical density of region lengths from the training set. Furthermore, the hidden semi-Markov model was enriched with n-grams representing signal peptide cleavage sites.

The second model is a simple probabilistic approach in which no association between amino acids was assumed, and probability of amino acids groups occurrence was determined by their frequencies in mature proteins.

@@

Eukaryotic protein sequences and their annotations were properly prepared according to the literature of the subject and downloaded from UniProt database release 2015\_06. The positive set contained 2~589 sequences with an experimentally confirmed signal peptide, its start and cleavage site. Sequences with more than one cleavage site were excluded from the final data set. The negative set comprised 152~272 sequences without any signal peptide annotation. Protein sequences with ambiguous symbols: X, J, Z and B were removed from the final sets. Proteins with selenocysteine (U) were also excluded from data set, because there are no records of signal peptides containing this amino acid.

@@

It is worth to note, that signalHsmm is not a first software to use amino acid encoding in signal peptide prediction. BLOMAP~\citep{maetschke2005blomap} also employed similar strategy, but considered only encodings based on the substitution matrices. Substitution matrices are computed using protein sequences and knowing how atypical signal peptides are, we had chosen different approach.

Instead amino acids were clustered using several criteria relevant for the architecture of signal peptide: hydrophobicity, frequency in alpha-helices, polarity and size. High hydrophobicity is a determinant of the h-region, the core of signal peptides. Alpha-helix, the secondary structure of the h-region, is probably induced by the positively charged n-region. High polarity as well as smaller size are important features of the residues in the cleavage site~\citep{1994palzkillselection}.

\begin{table}[ht]
\centering
\caption{Properties used in clusterization.} 
\begin{tabular}{ll}
  \toprule
Criterion name & Property name \\ 
  \midrule
Size & Size~\citep{dawson1972size} \\ 
   \rowcolor[gray]{0.85}Size & Molecular weight~\citep{fasman1976handbook}\\ 
  Size & Residue volume~\citep{1973goldsackcontribution} \\ 
   \rowcolor[gray]{0.85}Size & Bulkiness~\citep{1968zimmermanthe} \\ 
  Hydrophobicity & Normalized hydrophobicity scales for alpha-proteins~\citep{1992cidhydrophobicity} \\ 
   \rowcolor[gray]{0.85}Hydrophobicity & Consensus normalized hydrophobicity scale~\citep{1984eisenbergthreedimensional} \\ 
  Hydrophobicity & Hydropathy index~\citep{1982kytea} \\ 
   \rowcolor[gray]{0.85}Hydrophobicity & Surrounding hydrophobicity in alpha-helix~\citep{1980ponnuswamyhydrophobic} \\ 
  Polarity & Polarity~\citep{1974granthamamino} \\ 
   \rowcolor[gray]{0.85}Polarity & Mean polarity~\citep{1988radzickainfluences} \\ 
  Frequency in alpha-helices & Signal sequence helical potential~\citep{1982argosstructural} \\ 
   \rowcolor[gray]{0.85}Frequency in alpha-helices & Normalized frequency of N-terminal helix~\citep{chou1978prediction} \\ 
  Frequency in alpha-helices & Relative frequency in alpha-helix~\citep{1990prabhakaranthe} \\ 
   \bottomrule
\end{tabular}
\label{tab:aaprop}
\end{table}

We selected 13 properties from AAIndex database~\citep{2008kawashimaaaindex} (see Tab.~\ref{tab:aaprop}), each attributed to a single criterion. We established 96 permutations of the properties, where each permutation contains only one property associated with a given criterion. Considering all permutations of properties, we created 96 possible clusterings of amino acids using Euclidean distance and Ward's method. Further, we cut the clusterings to create four-group encodings. As expected, some encodings (31\%) were identical.

To compare encodings we performed a 5-fold cross-validation training a new instance of signalHsmm on every encoding. We created balanced data sets by subsampling a number of proteins without a signal peptide equal to the number of proteins with signal peptide.
The analysis was carried using also redundant encodings, but results are not shown (HSMM is a deterministic algorithm and predictions for identical groupings of amino acids were the same). The cross-validation was repeated 60 times. At first, we planned 160 repetitions to ensure with probability 0.95, that every protein without signal peptide was in the learning set, but very small variance of performance measures (for example see Tab.~\ref{tab:perfmeas}) allowed us to relax this condition wihout sacrificing credibility of cross-validation.

\subsection*{Hidden semi-Markov model}
Hidden semi-Markov model (HSMM) is an extension of hidden Markov model (HMM).
As such, it proved to yield better performance in many fields of bioinformatics including
gene recognition~\citep{Pachter02applicationsof}, protein secondary structure prediction~\citep{16571137}
or automatic detection of heartbeats~\citep{7043102}.
Let us first briefly describe the idea that lies behind HMM. 
Suppose you have a sequence of observations e.g. amino acids, and you are interested in understanding underlying cause of their occurrence. 
HMM aims to answer that question by assuming a specific, yet flexible, structure of the problem.
HMM consists of two stochastic processes. First is discrete Markov chain $X_{t=1}^T$ on the set of so called hidden states $\{S_1, \dots S_n\}$.
They are "the cause" of the observations. At every step $t$, hidden state might change according to transition matrix
$A= (a)_{i,j=1}^n$, where $a_{i,j} = \mathcal{P}(X_{t+1} = S_j | X_t = S_i)$. In our application hidden states are signal peptide regions.
Second process $E_{t=1}^T$ is observation process, defined on the set of possible observations $\{O_1, \dots, O_m\}$. 
Observations are assumed to occur independently. Their distribution depends only on hidden state which emits it. 
Probabilities are given by a matrix $B$, $b_{i,k} = \mathcal{P}(E_t = O_k | X_t = S_i)$.
In our case, observations are (degenerated) amino acids.
Main goal in signalHsmm is to find, for a given peptide, most probable regions boundaries. This is achieved with
Viterbi algorithm.
For a good reference on HMM see~\citep{1989rabinera}.

In regular HMM, hidden state duration -- number of observations emitted by the hidden state, has geometric distribution.
\citep{Durbin98biologicalsequence} shows how to extend it for some different distributions without significant increase in computational 
complexity. Similar ideas were used for signal peptide recognition, for example in~\citep{2004klla}. 
It is, however, still not flexible enough. Empirical regional length distributions (see Fig.~\ref{fig:reglen})
are difficult to capture in that way.

@@

(1,-2.8) -- (12,-2.8) node [branch2, midway, yshift=-1.5cm,color=black] {\textbf{Observations, $\sum_{i=1}^k d_i = T$}};

\end{tikzpicture}
\caption{General scheme of hidden semi-Markov model.}
\label{fig:hsmm}
\end{figure}

The model we use is Hidden semi-Markov Model(HSMM)~\citep{Yu2010215}. It extends HMM by allowing 
any given hidden state duration distribution. Diagram of HSMM is presented in Fig.~\ref{fig:hsmm}.
In addition to matrices $A$ and $B$, model is given by probabilities of duration length in hidden states.
$$\mathcal{P}(\text{duration in state} = d | \text{state is } S_i), \;\; i = 1, \dots, n, \;\; d = 1, \dots, D$$
where $D$ is maximum allowed duration.
As our datasets are of reasonable size, $D$ is small -- around 30, computational effort is not much 
higher than with the regular HMM.

The model we use has a very specific structure. Hidden states are signal peptide regions.
Almost all entries in transition matrix $A$ are zeros, because regions are sequential.
Possible transitions are depicted as arrows in Fig.~\ref{fig:signalHsmm}.
Probabilities of observations for hidden states and hidden states durations were estimated from training data.
The advantage of HSMM model is not only a better performance but also its straightforwardness.
Fig.~\ref{fig:signalHsmm} is easy to interpret for a researcher without any mathematical background.

@@

\subsection*{n-gram extension}

Hidden semi-Markov models may be flexibly enhanced by adding additional hidden states. To improve our model, we added few supplementary states representing specific motives that might occur in proximity of cleavage site. The structure of cleavage sites, more conservative  than other parts of signal peptide~\citep{2004hillerpredisi}, may be mirrored by n-grams (k-mers), short vectors of $n$ characters derived from input sequences.

Using the biogram software~\citep{biogramPackage} we extracted n-grams from cleavage sites of signal peptides. The analyzed sequences were already encoded using amino acid classification providing the best sensitivity of a general model (see Tab.\ref{tab:best}. Selected n-grams, representing less common cleavage site motifs, were included in the HSMM model as the alternative paths at the end of c-region. The architecture of extended HSMM can be seen in Fig.~\ref{fig:ngramext}.

>>>>

\title{SignalHsmm - a novel semi-Markov model of eukaryotic signal peptides}

@@

\keywords{eukaryote; hidden semi-Markov models; n-gram; signal peptide prediction}

\begin{abstract}
Short stretches of amino acid residues called targeting signals are involved in the proper localization of proteins in a cell, which is essential to maintain their desired function. Ones of them are signal peptides, which are responsible for targeting of proteins to endomembrane system and their export outside the cell. Proteins equipped with signal peptides play crucial roles in metabolism, maintenance of tissue structure, immune response and regulation of other organismal functions. Studies of many protein sequences indicate a great variability of signal peptides, which was not considered in the current software for their prediction. Therefore, we designed a new, more universal probabilistic model for eukaryotic signal peptides, which includes knowledge about their organization, amino acid composition and variation. The proposed approach is based on hidden semi-Markov models (HSMMs) and use intrinsic knowledge about signal peptides. The big advantage of the algorithm is its extensibility. Using the n-grams (k-mers), we showed that the general model can yield not only better results than other software but also more information about features of signal peptides. Our model showed the largest AUC = 0.97 in comparison to other signal peptide predictor even after training on very small data sets. Moreover, it appeared very stable regardless of types of learning data sets. Therefore, our model does not need to be permanently retrained with the continuous expansion of sequence databases. It should be emphasized that our model described signal peptides from medically significant malaria parasites \textsl{Plasmodium} and their relatives more accurately (with AUC = 0.92) than popular programs (0.84). The web-server of signalHsmm is available at \url{http://smorfland.uni.wroc.pl/signalhsmm}.

@@

\subsection*{Roles and features of signal peptides}

Proteins of eukaryotes are encoded in nuclear genomes and are synthesized in ribosomes located in the cytosol or bounded by the endoplasmic reticulum. After translation, proteins are targeted to specific subcellular compartments or exported outside the cell. The proper localization of proteins is essential to perform their desired function. Information about the protein destination is included within the very protein in short stretches of amino acid residues called targeting or sorting signals. Ones of them are signal peptides, which are located at the N-terminus of proteins.

Signal peptides are responsible for targeting of proteins via the Sec61 translocation channel~\citep{2007rapoportprotein} to endomembrane system, which includes endoplasmic reticulum and Golgi apparatus. Such proteins can stay inside of these compartments, can be inserted into cellular membranes or exported outside the cell. Proteins equipped with signal peptides play crucial roles in metabolism ($\beta$ galactosidase, pepsins)~\citep{1991hofmannmutations}, maintenance of tissue structure (collagen)~\citep{2001chanaberrant}, immune response (interferons, interleukins)~\citep{2005zhangalteration} and regulation of other organismal functions (prolactin, glucagon)~\citep{2010huangrole}. Moreover, passing proteins through the endomembrane system is important for their correct folding and posttranslational modification such as glycosylation and phosphorylation.

Despite the low sequence homology between signal peptides~\citep{1999ladungaphysean}, some general architecture were proposed~\citep{1994izardsignal, 2013vossmechanism} - Fig. \ref{fig:sparch}. It is assumed that signal peptides start with a positively charged sequence of amino acid residues, called the n-region with the length of about 5-8 residues. They probably enforce a proper topology on the polypeptide during its translocation through membrane based on the positive-inside rule~\citep{1988vonheijnetopogenic}. The first region is followed by a stretch of hydrophobic amino acids (h-region) with the length of about 8-12 residues. It constitutes a core region of signal peptide and usually forms $\alpha$-helix. The third part of a signal peptide, usually 6 residues long, is a polar, but uncharged c-region ended with a cleavage site, in which the signal peptidase cleaves the signal peptide, during or after translocation of the protein into the endoplasmic reticulum ~\citep{2002paetzelsignal}. The cleavage site is characterized by a variable amino acid composition. It typically contains small and neutral residues at -3 and -1 positions~\citep{1994palzkillselection}. This site is, however, absent from some membrane proteins in which the first transmembrane domain acts both as a signal peptide and signal anchor~\citep{1988szczesnaskorupapositive}. The amino acid composition and the length of these regions vary between signal peptides, which influences the efficiency of protein secretion~\citep{2006hegdethe}.

\begin{figure}[ht]\centering
\includegraphics[width=0.55\textwidth]{figures/SP.png}
\caption{The organization of typical signal peptide. The figure is not drawn in scale.}
\label{fig:sparch}
\end{figure}

Some data indicate that signal peptides may be universal. It was found, for example, that even bacterial signal peptides targeted correctly transgenic proteins to plant~\citep{2009moellera} or mammalian secretory systems~\citep{2014naganoestablishment}. On the other hand, signal peptides show great variation and the description presented above (Fig. \ref{fig:sparch}) refers to the most ‘typical’ signal peptides. There are  exceptionally long signal peptides, which fulfill more sophisticated roles~\citep{2009hissarchitecture}. The fragment of signal peptide from preprolactin takes part in the regulation of prolactin secretion, whereas signal peptides of MHC class I inhibit activity of NK cells. Signal peptides of viral origin are involved in the immune evasion or viral life cycle~\citep{2000kappposttargeting}. The signal peptide from midkine contributing to the tumor progression contains epitopes recognized by CD4+ T cells~\citep{2013kerzerhothe}. The functional significance of these targeting signals makes that the prediction of signal peptide-containing proteins is also an important step in the drug development~\citep{2005zhangalteration, 2012netoadeimproving, 2010moellerwetmilling}.

\subsection*{Software predicting signal peptides}

Although many experimental methods determining the subcellular localization of proteins were devised, they are time consuming and laborious. Therefore, signal peptides became the subject of many computational programs to their prediction. Many software incorporates ‘black-box’ models, such as: neural networks~\citep{2011petersensignalp}, support vector machines~\citep{2014zhangprediction}, Bayesian networks~\citep{2012zhengsignalbnf} or k-nearest neighbours~\citep{2007shensignall}. However, these models do not provide direct biological information about organization of signal peptides and are not able to predict properly atypical signal peptides. Although there are programs that do not share the innate flaws of ‘black-box’ models, they also demand an improvement. Some of them are based on position matrices or their variants~\citep{2014zhangprediction, 2004hillerpredisi}. Others (Phobius, Philius and SignalP 3.0) use hidden Markov models (HMMs)~\citep{2004klla, 2008reynoldstransmembrane, 2004bendtsenimproved}, which try to reflect regions of signal peptides without realizing limitations of their probabilistic frameworks. The used HMMs imply a geometric distribution for duration of regions' length. We verified the distribution for regions from the first work utilizing HMMs in prediction of signal peptides~\citep{1998nielsenprediction} and found that the the length distribution for every region was other than geometric (Fig. \ref{fig:reglen}). Moreover, the commonly used rigid scheme of signal peptide's organization (Fig. \ref{fig:sparch}) does not describe extremely long or short peptides. Theoretically, HMMs that describe the atypical signal peptides could be developed to consider also the unusual structures but such probabilistic frameworks have not still been implemented.

\begin{figure}[ht]\centering
\includegraphics[width=0.95\textwidth]{figures/reglen.eps}
\caption{Distribution of lengths of signal peptides (A) and their regions (B) expressed in the number of amino acid residues. The data was extracted from 2~589 signal peptide sequences derived from UniProt database (see \textbf{Data selection} in \textbf{Methods}).}
\label{fig:reglen}
\end{figure}

All programs used in signal peptide recognition are trained on real protein sequences. Therefore, they succeed in the recognition of peptides similar to those in the learning set but fail in the case of artificial signal peptides. Such peptides are designed to increase effectiveness of protein secretion~\citep{2010futatsumorisugaisignal}. They are especially important in industrial applications to increase yield of proteins. Therefore, only explicit knowledge about the organization of signal peptides allows creating sequences that will be the most efficient in the export of proteins~\citep{2013ngengineering}. Signal peptides have also an important application in gene therapy. Mimicking the natural mechanism of protein export, artificial signal peptides with tumor epitopes increase the antitumor immune response~\citep{2003heenhanced}. Such epitopes must be properly inserted into a signal peptide without decreasing its secretion properties through disruption of the regional structure. Instead of time-consuming and expensive laboratory experiments, it would be very useful to survey in silico many artificial peptides to select the ones that would fulfill the designed role.

Majority of the signal peptide predicting software use the orthogonal encoding of amino acids, in which a vector of 20 digits represents every amino acid. This method of encoding, however, does not regard relationships between amino acids and differences in their physicochemical properties. This is disadvantage of such signal peptides' models because their regions are in fact characterized by specific features of amino acid residues and not by the simple occurrence of specific amino acids. In addition to this, such sparse encoding enforces larger data sets, which hinders their management and analysis~\citep{2002linamino}. Therefore, we elaborated a new approach based on hidden semi-Markov models using grouping of amino acids into physicochemical groups characteristic of signal peptides. The new methods proved better in comparison to the current software.

@@

Since the functionality of signal peptides depends the physicochemical properties of residues in a given region, we clustered amino acids into several groups based on their characteristics. The pre-processed sequences were further analyzed by the heuristic algorithm, which determines borders between three characteristic signal peptide regions~\cite{1998nielsenprediction}. We refined some region recognition criteria to attune the algorithm to less typical signal peptides. Next, two models were trained to recognize proteins with and without a signal peptide. The first one was a hidden semi-Markov model, in which each of three signal peptide regions was represented by a different hidden state. The additional fourth hidden state represented a mature protein. Each state was described by its frequencies of amino acid groups. The distribution of hidden states durations, i.e. the number of amino acids, was based on the empirical distribution of region lengths from the training set. Furthermore, the hidden semi-Markov model was enriched with n-grams representing signal peptide cleavage sites. The second model was a simple probabilistic approach in which no association between amino acids was assumed, and probability of amino acids groups occurrence was determined by their frequencies in mature proteins.

@@

Eukaryotic protein sequences and their annotations were properly prepared according to the literature of the subject and downloaded from UniProt database release 2015\_06. The positive set contained 2~589 sequences with an experimentally confirmed signal peptide, its start and cleavage site. Sequences with more than one cleavage site were excluded from the final data set. The negative set comprised 152~272 sequences without any signal peptide annotation. Protein sequences with ambiguous symbols: X, J, Z, B and selenocysteine (U) were removed from the final sets.

@@

SignalHsmm is not the first software which uses amino acid encoding in signal peptide prediction. BLOMAP~\citep{maetschke2005blomap} also employed the similar strategy but considered only substitution matrices. However, we applied different approach. We clustered amino acids using four properties relevant for the architecture of signal peptide: their hydrophobicity, frequency in $\alpha$-helices, polarity and size. The high hydrophobicity is determinants of the h-region, whose $\alpha$-helix secondary structure is probably induced by the positively charged n-region. The high polarity as well as small size are important features of residues in the cleavage site~\citep{1994palzkillselection}.

\begin{table}[ht]
\small
\centering
\caption{Properties used in amino acid clusterization.} 
\begin{tabular}{ll}
  \toprule
Property name & Amino acid scale and its reference \\ 
  \midrule
Size & Size~\citep{dawson1972size} \\ 
   \rowcolor[gray]{0.85}Size & Molecular weight~\citep{fasman1976handbook}\\ 
  Size & Residue volume~\citep{1973goldsackcontribution} \\ 
   \rowcolor[gray]{0.85}Size & Bulkiness~\citep{1968zimmermanthe} \\ 
  Hydrophobicity & Normalized hydrophobicity scales for $\alpha$-proteins~\citep{1992cidhydrophobicity} \\ 
   \rowcolor[gray]{0.85}Hydrophobicity & Consensus normalized hydrophobicity scale~\citep{1984eisenbergthreedimensional} \\ 
  Hydrophobicity & Hydropathy index~\citep{1982kytea} \\ 
   \rowcolor[gray]{0.85}Hydrophobicity & Surrounding hydrophobicity in $\alpha$-helix~\citep{1980ponnuswamyhydrophobic} \\ 
  Polarity & Polarity~\citep{1974granthamamino} \\ 
   \rowcolor[gray]{0.85}Polarity & Mean polarity~\citep{1988radzickainfluences} \\ 
  Frequency in $\alpha$-helices & Signal sequence helical potential~\citep{1982argosstructural} \\ 
   \rowcolor[gray]{0.85}Frequency in $\alpha$-helices & Normalized frequency of N-terminal helix~\citep{chou1978prediction} \\ 
  Frequency in $\alpha$-helices & Relative frequency in $\alpha$-helix~\citep{1990prabhakaranthe} \\ 
   \bottomrule
\end{tabular}
\label{tab:aaprop}
\end{table}

We considered in total 13 amino acid scales from AAIndex database~\citep{2008kawashimaaaindex} (Tab.~\ref{tab:aaprop}). We selected one scale per property and carried out all possible 96 permutations of them. Based on that, we created 96 possible clusterings of amino acids using Euclidean distance and Ward's method. Next, we cut the clusterings to create four group of amino acids. In 31\% of cases, the groupings were identical. To compare the usefulness of encodings, we performed a 5-fold cross-validation training a new instance of signalHsmm on every encoding. We created balanced data sets by subsampling proteins without a signal peptide to equal the number of proteins with a signal peptide. The cross-validation was repeated 60 times. At first, we planned 160 repetitions to ensure that every protein without signal peptide was in the learning set with the probability of 0.95, but a very small variance of performance measures (for example see Tab.~\ref{tab:perfmeas}) for smaller number of repetitions allowed us to relax this condition without decreasing credibility of cross-validation.

\subsection*{Hidden semi-Markov model}
Hidden semi-Markov model (HSMM) is an extension of hidden Markov model (HMM). 
Let us briefly describe the idea of the HMM. 
Suppose we have a sequence of observations, e.g. amino acids, and we are interested in understanding an underlying cause of their occurrence. 
HMM aims to answer that question assuming a specific and flexible structure of the problem.
HMM consists of two stochastic processes. The first is a discrete Markov chain $X_{t=1}^T$ on the set of so called hidden states $\{S_1, \dots S_n\}$.
They are "the cause" of the observations. At every step $t$, hidden state might change according to a transition matrix
$A= (a)_{i,j=1}^n$, where $a_{i,j} = \mathcal{P}(X_{t+1} = S_j | X_t = S_i)$. In our application, hidden states are signal peptide regions.
Th second process $E_{t=1}^T$ is an observation process defined on the set of possible observations $\{O_1, \dots, O_m\}$, which are assumed to occur independently. Their distribution depends only on the hidden state which emits it. 
Probabilities are given by a matrix $B$, $b_{i,k} = \mathcal{P}(E_t = O_k | X_t = S_i)$.
In our case, observations are (encoded) amino acids.
The main goal in signalHsmm is to find for a given peptide most probable regions boundaries. This is achieved with Viterbi algorithm.
For a good reference on HMM, see~\citep{1989rabinera}.

In the regular HMM, the hidden state duration, i.e. the number of observations emitted by the hidden state, has a geometric distribution. \citep{Durbin98biologicalsequence} showed how to extend it for different distributions without significant increase in computational 
complexity. Similar ideas were used for signal peptide recognition, for example by~\citep{2004klla}. 
However, it is still not flexible enough because the empirical regional length distributions (see Fig.~\ref{fig:reglen})
are difficult to capture in this way.

@@

\draw [color=red!30!black!50, line width=1.5mm, decorate,decoration={brace,amplitude=15pt,mirror,raise=10pt},yshift=-10pt]
(1,-2.6) -- (12,-2.6) node [branch2, midway, yshift=-1.3cm,color=black] {\textbf{Observations, $\sum_{i=1}^k d_i = T$}};
\end{tikzpicture}
\caption{General scheme of hidden semi-Markov model.}
\label{fig:hsmm}
\end{figure}

The model we used is in fact Hidden semi-Markov Model(HSMM)~\citep{Yu2010215}. It extends HMM by allowing any given hidden state duration distribution (Fig.~\ref{fig:hsmm}).
In addition to matrices $A$ and $B$, the model is given by probabilities of duration length in hidden states.
$$\mathcal{P}(\text{duration in state} = d | \text{state is } S_i), \;\; i = 1, \dots, n, \;\; d = 1, \dots, D$$
where $D$ is the maximum allowed duration.
As our datasets are of reasonable size, $D$ is small -- around 30, computational effort is not much 
higher than in the regular HMM.

Our model has a very specific structure. Ihe hidden states represent signal peptide regions.
Almost all entries in the transition matrix $A$ are zeros because regions are sequential.
Possible transitions are depicted as arrows in Fig.~\ref{fig:signalHsmm}.
Probabilities of observations for the hidden states and hidden states durations were estimated from training data.
The advantage of HSMM model is not only a better performance but also its straightforwardness.
Fig.~\ref{fig:signalHsmm} is easy to interpret for a researcher without any mathematical background.

@@

\subsection*{Extension of n-grams}

Hidden semi-Markov models may be flexibly enhanced by adding additional hidden states. To improve our model, we added few supplementary states representing specific motives that might occur in the proximity of cleavage site. The structure of cleavage sites, more conserved than other parts of signal peptide~\citep{2004hillerpredisi}, may be mirrored by n-grams (k-mers), short vectors of $n$ characters derived from input sequences. Using the biogram software~\citep{biogramPackage}, we extracted n-grams from cleavage sites of signal peptides. The analyzed sequences were already encoded using amino acid classification providing the best sensitivity of the general model (see Tab.\ref{tab:best}. Selected n-grams representing less common cleavage site motifs were included in the HSMM model as alternative paths at the end of c-region (Fig.~\ref{fig:ngramext}).
